# Functions {#functions}

This chapter will explore how to use **functions** in R to perform advanced capabilities and actually ask questions about data. After considering a function in an abstract sense, it will discuss using built-in R functions, accessing additional functions by loading R packages, and writing your own functions.


## What are Functions?

In a broad sense, a **function** is a named sequence of instructions
(lines of code) that you may want to perform one or more times
throughout a program. They provide a way of _encapsulating_ multiple
instructions into a single "unit" that can be used in a variety of
different contexts. So rather than needing to repeatedly write down all
the individual instructions for "make a sandwich" every time you're
hungry, you can define a `make_sandwich()` function once and then just
**call** (execute) that function when you want to perform those steps.
Typically, functions also accept _inputs_ so you can make the things
slightly differently from time to time.  For instance, sometimes you may
want to call `make_sandwich(cheese)` while another time `make_sandwich(chicken)`.

In addition to grouping instructions, functions in programming languages
also allow to model the mathematical definition of function,
i.e. perform certain operations on a number of inputs that lead to an
_output_.  For instance, look at the
function `max()` that finds the largest value among numbers:

```{r, eval=FALSE}
max(1,2,3)  # 3
```
The inputs are numbers `1`, `2`, and `3` in parenthesis, usually called
**arguments** or **parameters**, and we say that these arguments are
**passed** to a function (like a football). We say that a function then
**returns** a value, number "3" in this example, which we can either
print or assign to a variable and use later.

Finally, the functions may also have **side effects**.  An example case
is the `cat()` function that just prints it's arguments.  For instance,
in case of the following line of code
```{r, eval=FALSE}
cat("The answer is", 1+1, "\n")
```
we call function `cat()` with three arguments: `"The answer is"`, `2`
(note: `1+1` will be evaluated to 2),
and `"\n"` (the line break symbol).  However, here we may not care about the return value but in the side effect&mdash;the message being printed on the screen.


## How to Use Functions {#how-to-use-functions}

R functions are referred to by name (technically, they are values like
any other variable, just not atomic values). As in many programming
languages, we **call** a function by writing the name of the function
followed immediately (no space) by parentheses `()`.  Sometimes this is
enough, for instance
```{r, eval=FALSE}
Sys.Date()
```
gives us the current date and that's it.

But often we want the function to do something with our inputs.  In this
case we put the **arguments** (inputs) inside the parenthesis, separated
by commas (**`,`**). Thus computer functions look just like
multi-variable mathematical functions, although usually with fancier names than `f()`.

```r
# call the sqrt() function, passing it 25 as an argument
sqrt(25)  # 5, square root of 25

# count number of characters in "Hello world" as an argument
nchar("Hello world")  # 11, note: space is a character too

# call the min() function, pass it 1, 6/8, AND 4/3 as arguments
# this is an example of a function that takes multiple args
min(1, 6/8, 4/3)  # 0.75, (6/8 is the smallest value)
```

To keep functions and ordinary variables distinct, we include empty parentheses `()` when referring to a function by name. This does not mean that the function takes no arguments, it is just a useful shorthand for indicating that something is a function.

<p class="alert alert-info">**Note:** You always need to supply the
parenthesis if you want to _call_ the function (force it to do what it
is supposed to do).  If you leave the parenthesis out, you get the
function definition printed on screen instead.  So `cat()` is actually a _function call_ while `cat` is the function.  You can see that it is a function if you just print it as `print(cat)`.
However, we ignore this distinction here.
</p>

If you call any of these functions interactively, R will display the **returned value** (the output) in the console. However, the computer is not able to "read" what is written in the console&mdash;that's for humans to view! If you want the computer to be able to _use_ a returned value, you will need to give that value a name so that the computer can refer to it. That is, you need to store the returned value in a variable:

```r
# store min value in smallest.number variable
smallest_number <- min(1, 6/8, 4/3)

# we can then use the variable as normal, such as for a comparison
min_is_big <- smallest_number > 1  # FALSE

# we can also use functions directly when doing computations
phi <- .5 + sqrt(5)/2  # 1.618...

# we can even pass the result of a function as an argument to another!
# watch out for where the parentheses close!
print(min(1.5, sqrt(3)))  # prints 1.5
```

- In the last example, the resulting _value_ of the "inner" function
  (e.g., `sqrt()`) is immediately used as an argument for the middle
  function (i.e. `min()`), value of which is fed in turn to the outer
  function `print()`.  Because that value is used immediately, we don't
  have to assign it a separate variable name. It is known as an
  **anonymous variable**.
- note also that in the last example, we are solely interested in the
  side effect of the `print()` function.  It also returns it's argument
  (`min(1.5, sqrt(3))` here) but we do not store it in a variable.

R functions take two types of arguments: **positional arguments** and
**named arguments**: the function has to know how to treat each of it's
arguments.  For instance, we can round number _e_ to 3
digits by `round(2.718282, 3)`.  But in order to do this, the `round()` function must know that `2.718282` is the number
and `3` is the requested number of digits, and not the other way around.  It understands this because 
it requires the number to be the first argument, and digits
the second argument.  This approach works well in case of known small
number of inputs.  However, this is not an option for functions with
**variable number of arguments**, such as `cat()`.  `cat()` just prints
out all of it's (potentially a large number of) inputs, except a limited number of special named
arguments.  One of these is `sep`, the string to be placed between the other pieces
of output (by default just a space is printed).  Note the difference in output
between 
```{r, eval=FALSE}
cat(1, 2, "-", "\n")  # 1 2 - 
cat(1, 2, sep="-", "\n")  # 1-2-
```
In the first case `cat()` prints `1`, `2`, `"-"`, and the line break
`"\n"`, all separated by a space.  In the second case the name `sep`
ensures that `"-"` is not
printed out but instead treated as the separator between `1`, `2` and `"\n"`.


## Built-in R Functions

As you have likely noticed, R comes with a variety of functions that are built into the language. In the above example, we used the `print()` function to print a value to the console, the `min()` function to find the smallest number among the arguments, and the `sqrt()` function to take the square root of a number. Here is a _very_ limited list of functions you can experiment with (or see a few more [here](http://www.statmethods.net/management/functions.html)).

<!-- TODO: Fix the formatting for this? -->

| Function Name | Description | Example |
| :------------- | :---------------------- | :------------------------------- |
| `sum(a,b,...)`  | Calculates the sum of all input values  | `sum(1, 5)` returns `6`|
| `round(x,digits)` | Rounds the first argument to the given number of digits | `round(3.1415, 3)` returns `3.142` |
| `toupper(str)`  | Returns the characters in uppercase  | `toupper("hi there")` returns `"HI THERE"` |
| `paste(a,b,...)` | _Concatenate_ (combine) characters into one value | `paste("hi", "there")` returns `"hi there"` |
| `nchar(str)` | Counts the number of characters in a string | `nchar("hi there")` returns `8` (space is a character!) |
| `c(a,b,...)` | _Concatenate_ (combine) multiple items into a _vector_ (see [chapter 7](#vectors)) | `c(1, 2)` returns `1, 2` |
| `seq(a,b)`  | Return a sequence of numbers from a to b  | `seq(1, 5)` returns `1, 2, 3, 4, 5` |


To learn more about any individual function, look them up in the R documentation by using `?FunctionName` account as described in the previous chapter.

<!-- TODO: Make this a link to the chapter?? Why won't it work? -->

<p class="alert alert-info">"Knowing" how to program in a language is to some extent simply "knowing" what provided functions are available in that language. Thus you should look around and become familiar with these functions... but **do not** feel that you need to memorize them! It's enough to simply be aware "oh yeah, there was a function that sums up numbers", and then be able to look up the name and argument for that function.</p>


## Loading Functions

Although R comes with lots of built-in functions, you can always use
more! **Packages** (also known as **libraries**) are additional sets of
R functions (and data and variables) that are written and published by the R community. Because
many R users encounter the same data management/analysis challenges,
programmers are able to use these libraries and thus benefit from the
work of others (this is the amazing thing about the open-source
community&mdash;people solve problems and then make those solutions
available to others).  Popular packages include _dplyr_ for
manipulating data, _ggplot2_ for visualizations, and _data.table_ for
handling large datasets.

Most of the R packages **do not** ship with the R
software by default, and need to be downloaded (once) and then loaded
into your interpreter's environment (each time you wish to use
them). While this may seem cumbersome, it is a necessary trade-off
between speed and size.  R software would be huge and slow if it would
include _all_ available packages.

Luckily, it is quite simple to install and load R packages from within
R. To do so, you'll need to use the _built-in_ R functions
`install.packages` and `library`. Below is an example of installing and
loading the `stringr` package (which contains many handy functions for working with character strings):

```r
# Install the `stringr` package. Only needs to be done once on your machine
install.packages("stringr")
```
<p class="alert alert-warning">
We stress here that you need to install each package only once per computer.
As installation may be slow and and resource-demanding, you **should not
do it repeatedly inside your script!**.  Even more, if your script is
also run by other users on their computers, you should get their
explicit consent before installing additional software for them!
The easiest remedy is to solely rely on manual installation.
</p>

Exactly the same syntax&mdash;`install.packages("stringr")`&mdash;is
also used for re-installing it.  You may want to re-install it if a
newer version comes out, of if you upgrade your R and receive warnings 
about the package being built under a previous version of R.
  
After installation, the easiest way to get access to the functions is by
_loading the package_:
```{r, eval=FALSE}
# Load the package (make stringr() functions available in this R session/program)
library("stringr") # quotes optional here
```
This makes all functions in the _stringr_ package available for R (see [the documentation](https://cran.r-project.org/web/packages/stringr/stringr.pdf) for a list of functions included with the `stringr` library).  For
instance, if we want to pad the word "justice" from left with tildes to create a
width-10 string, we can do
```{r, eval=FALSE}
str_pad("justice", 10, "left", "~")  # "~~~justice"
```
We can use `str_pad` function without any additional hassle because `library()` command made it
available.

This is an easy and popular approach.  However, what happens if more than one package call a function by the same name?  For
instance, many packages implement function `filter()`.  In this case
the more recent package will _mask_ the function as defined by the
previous package.  You will also see related warnings when you load the
library.  In case you want to use a masked
function you can write something like `package::function()` in order to
call it.  For instance, we can do the example above with
```{r, eval=FALSE}
stringr::str_pad("justice", 10, "left", "~")  # "~~~justice"
```
This approach&mdash;specifying _namespace_ in front of the
function&mdash;ensures we access the function in the right package.  If
we call all functions in this way, we don't even need to load the
package with `library()` command.  This is the preferred approach if you
only need few functions from a large library.



## Writing Functions
Even more exciting than loading other peoples' functions is writing your
own. Any time you have a task that you may repeat throughout a
script&mdash;or sometimes when you just want to organize your code
better&mdash;it's a good practice to write a function to perform that task. This will limit repetition and reduce the likelihood of errors as well as make things easier to read and understand (and thus identify flaws in your analysis).

Functions are values like numbers and characters, so we use the _assignment
operator_ (**`<-`**) to store a function into a variable.  
<p class="alert alert-info">
Although "values" (more precisely _objects_), functions are not [atomic
objects](#basic-data-strucures).  But they can still be stored and
manipulated in many ways like other objects.
</p>

We explain how to write functions using an example.  Let's create a
function `fullName` that takes first name and last name as strings, and returns
full name, i.e. a single string made of the these names.
```r
fullName <- function(first, last) {
  full <- paste(first, last)
  return(full)
}
```
This is not particularly interesting function--it just concatenates
two strings--but it helps us to explain all the major building blocks
of functions.

Function definition contains several important pieces:

- The value assigned to the function name (here `fullName`) uses the
    syntax `function(....)` to indicate that you are creating a function
    (and not a number or character string).
	
- We usually want to feed the function with certain
    inputs, here the first and last name.
	These inputs must be referred to somehow inside of your function.
    You list these names in the parentheses in the `function(....)`
    declaration.  These are called **formal arguments**.  Formal arguments _will
    contain_ the values passed in when calling the function (called
    **actual arguments**).  For example, when we
    call `fullName("Alice", "Kim")`, the value of the first
    actual argument (`"Alice"`) will be assigned to the first formal argument
    (`first`), and the value of the second actual argument (`"Kim"`) will
    be assigned to the second formal argument (`last`).  Inside the
    function's body, both of these formal arguments behave exactly as
    ordinary variables with values "Alice" and "Kim" respectively.

	We could have made the formal argument names anything
    we wanted (`name_first`, `xyz`, etc.), just as long as we then use
    _these formal argument names_ to refer to the argument while inside
    the function. 
	
	The formal argument names are only valid within the function's
    body (inside of the curly braces).
	Variables `first`, `last`, and `full` only exist within this
    particular function, if you try to access those outside of the
    function, you get and error like `Error: object 'first' not found`.

- **Body**: The body of the function is a **block** of code that falls
  between curly braces **`{}`** (a "block" is represented by curly
  braces surrounding code statements).  The opening brace `{` must
  follow the `function(....)` declaration, the closing `}` will
  complete the function--whatever follows the closing brace is not
  part of the function any more.
  It is a tradition to put the opening `{` immediately after the arguments list, and the closing `}` on its own line.

    The function body specifies all the instructions (lines of code)
    that your function will perform. A function can contain as many
    lines of code as you want&mdash;you'll usually want more than 1 to
    make it worth while, but if you have more than 20 you might want to
    break it up into separate functions. You can use the formal
    arguments here exactly as you use any other variables.  You can
    also
	create new variables,
    call other functions, you can even declare functions inside
    functions... basically any code that you would write outside of a
    function can be written inside of one as well!
	
	But remember that all the variables you create in the function body are **local
	variables**.  These are only visible from within the function and
	"will be forgotten" as soon as you return from the function.  However,
	variables defined outside of the function are still visible from
	within. 

- **Return value** is what your function produces.  You can specify this
    by calling the `return()` function and passing it the value that
    you wish _your function_ to return.  It is typically the last line
    of the function.  Note that even though we returned a variable
    called `full`, that variable was _local_ to the function and so
    doesn't exist outside of it; thus we have to store the returned
    value into another variable if we want to use it later (for
    instance, as
    `name <- fullName("Alice", "Kim")`).

After we have defined the function, we can _call_ (execute) it.
We call a function we defined exactly in the same way we
call built-in functions.
For instance, we may call it as

```r
fullName("Alice", "Kim")
                           # will return "Alice Kim"
```
When we do so, R will take the _actual
arguments_ we passed in (here `"Alice"` and `"Kim"`) and assign these to
the _formal arguments_ (here `first` and `last`).
Then it executes each line of code in the
_function body_ one at a time. When it gets to the `return()` call, it
will end the function and return the given value, which can then be
assigned to a different variable outside of the function. 

`return()` statement is usually unnecessary as R implicitly returns
the last value it evaluated anyway.  So we may shorten the function
definition into
```{r, eval=FALSE}
fullName <- function(first, last) {
  full <- paste(first, last)
}
```
or even not store the concatenated names into `full`:
```{r, eval=FALSE}
fullName <- function(first, last) {
   paste(first, last)
}
```
The last evaluation was concatenating the first and last name, and
hence the full name will be implicitly returned.


## Returning values versus producing output

Beginners have often difficulties seeing difference between functions
that _produce output_ (i.e. print on screen, see more in 
[Producing output](#producing-output-cat-print)) and that _return a value_
(that will be automatically printed on screen).  For instance,
consider two functions that compute minutes in day:
```{r}
minutesDay1 <- function() {
   mid <- 24*60
   cat(mid, "\n")  # output the numbers
}
minutesDay2 <- function() {
   mid <- 24*60
   return(mid)  # return the number
                # 'return' is optional in R
}
```
When we call the former function on console, we get
```{r}
minutesDay1()
```
and when we call the latter, we get
```{r}
minutesDay2()
```
The outputs appear almost the same.  However, behind the
scenes (well, behind the printed output) there are important
differences:

* The first function produces output, i.e. it always prints the
  number 1440.  The second function does not produce any output.  The
  line `[1] 1440` we see when we call it is created by R console that
  automatically prints the last result.  We can see the difference
  when we, for instance, assign the result to a variable:
```{r}
mid1 <- minutesDay1()
mid2 <- minutesDay2()
```
  `minutesDay1()` still produces the output but `minutesDay1()` is now
  silent.  If the result is assigned to a variable, it will not be
  automatically printed.
  
* The second function _returns_ the value that can be assigned to a
  variable (`mid2` in this example).  The first function returns
  whatever its last statement, `cat` returns.  This appears to be a
  special empty symbol `NULL`:
```{r}
mid1
mid2
```
  We can see `mid1` variable is empty while `mid2` contains the
  expected value.

So the first example outputs the result but does not return it, and
the second function does not output it but returns the value.
One can also create a function that does both (or
neither).  But which approach is better?

Obviously, it depends on on what you are doing.  In practice it is
common to have two types of functions--one type only outputs and does
not compute anything, and the other type only computes and returns,
but leaves output for the dedicated output functions.

<p class="tbd">An exercise</p>


## Conditional Statements
Functions are one way to organize and control the flow of execution
(e.g., what lines of code get run in what order).  There are other
ways.  In all languages, we can specify that different instructions will
be run based on a different set of conditions.  These are **Conditional
statements** which specify which chunks of code will run
depending on which conditions are true.  This is valuable both within
and outside of functions.

In an abstract sense, an conditional statement is saying:

```
IF something is true
  do some lines of code
OTHERWISE
  do some other lines of code
```

In R, we write these conditional statements using the keywords **`if`** and **`else`** and the following syntax:

```r
if (condition) {
  # lines of code to run if condition is TRUE
} else {
  # lines of code to run if condition is FALSE
}
```

(Note that the the `else` needs to be on the same line as the closing `}` of the `if` block. It is also possible to omit the `else` and its block).

The `condition` can be any variable or expression that resolves to a logical value (`TRUE` or `FALSE`). Thus both of the conditional statements below are valid:

```r
porridge_temp <- 115 # in degrees F
if (porridge_temp > 120) {
  print("This porridge is too hot!")
}

too_cold <- porridge_temp < 70
# a logical value
if (too_cold) {
  print("This porridge is too cold!")
}
```

Note, we can extend the set of conditions evaluated using an `else if` statement. For example:

```r
# Function to determine if you should eat porridge
test_food_temp <- function(temp) {
  if (temp > 120) {
    status <- "This porridge is too hot!"
  } else if (temp < 70) {
    status <- "This porridge is too cold!"
  } else {
    status <- "This porridge is just right!"
  }
  return(status)
}
# Use function on different temperatures
test_food_temp(119)  # "This porridge is just right!"
test_food_temp(60)   # "This porridge is too cold!"
test_food_temp(150)  # "This porridge is too hot!"
```

See more about `if`, `else` and other related constructs in [Appendix C](#control-structures).

## Resources {-}
- [R Function Cheatsheet](https://cran.r-project.org/doc/contrib/Short-refcard.pdf)
- [User Defined R Functions](http://www.statmethods.net/management/userfunctions.html)
