# Exercise solutions {#solutions}

## Lists

### Vectors and lists {#solutions-vectors-vs-lists}

The vector will be
```{r}
c(1, 2:4, 5)
```
and the list
```{r}
list(1, 2:4, 5)
```
The printout clearly shows that in case of vector we end up with a
vector of 5 similar elements (just numbers).  But the list contains
three elements, the first and last are single numbers (well, more
precisely length-1 vectors), while the middle component is a length-3
vector.

As this example shows, one cannot easily print all list elements on a
single row as is the case with vectors.


### Print employee list {#solutions-print-employees}

First re-create the same persons:

```{r}
person <- list(name = "Ada", job = "Programmer", salary = 78000,
               union = TRUE)
person2 <- list("Ji", 123000, FALSE)
employees <- list(person, person2)
```
The printout looks like
```{r}
employees
```
We can see our two employees here, Ada (at first position) and Ji (at
second position).  All element names for Ada are preceded with `[[1]]`
and for Ji with `[[1]]`.  These indicate the corresponding positions.

Ada and Ji data itself is printed out slightly differently, reflecting
the fact that Ada's components have names while Ji's components do
not.  So Ada's components use `$name` tag and Ji's components use a
similar `[[1]]` positional tag.
